Creating optimal conditions for reproducible data analysis in R with `fertile`

### 2020-02-22 fertile.Rmd
updated citations + removed notes from document

As Goodman, Fanelli, and Ionnidis argue in their 2016 paper, *What does research reproducibilty mean?*, the language and conceptual framework of research reproducibility vary across the sciences. There are no clear standards agreed upon across fields.

The variety of recommendations and techniques for achieving reproducibility can be clearly seen Kitzes, Turek, and Deniz' book *The Practice of Reproducible Research*, a collection of case studies on reproducibility practices from across the data-intensive sciences. Although the book idenfies no consensus on exact standards of reproducibility, several common trends and principles emerge from the case studies:

@@

*A guide to reproducible code in ecology and evolution*, focused on data analysis in R, identifies similar important reproducibility components, reiterating the need to clearly labled, well-documented, and well-separated files. In addition, the paper recommends that a list of dependencies be published and version control be used.

@@

Much of the available literature focuses on file structure, organization, and naming, and **fertile`**s features are consistent with this. The *ideal* file structure is not agreed upon by academics, though there are some recommendations focused on R that are available. *Packaging data analytical work reproducibility using {R} (and friends)* by Marwick, Boettiger, and Mullen provides the framework for file structure that **fertile** is based on: a structure similar to that of an R package, with an `R` folder, as well as `data`, `data-raw`, `inst` and `vignettes`.

@@

As we have seen, many existing R packages focusing on reproducibility are designed in a way such that they can only be used prior to the start of a project, often for the purposes of creating an organized file structure or ordered file hierarchy. 
**fertile**, on the other hand, can be used at any stage in the data analysis workflow. It is designed to be used on data analyses organized as R Projects (i.e. folders containing a `.Rproj` file). Once an R Project is created, **fertile** can provide benefits at any time, both while coding as well as after the fact.
**fertile** operates in two modes: proactively (to prevent reproducibility mistakes from happening in the first place), and retroactively (analyzing code that has already been written for potential problems). 
`fertile` is available for download at [https://github.com/baumer-lab/fertile](https://github.com/baumer-lab/fertile)

@@

# The Importance of Ensuring Reproducibility

One reason for this is that reproducible research encourages transparency. Providing one's audience with well organized, reproducible code and results allows readers to understand the steps taken to generate findings from raw data and determine for themselves whether or not they believe the results to be reliable or trustworthy.

Additionally, reproducibility encourages collaboration and extended research, allows others to easily apply the methods used in one project to their own work with minimal effort. @bray2014five

Fertile is designed to make reproducibility simple, providing fast and easy methods to test an R project for reproducibility. The package is intended to be usable by introductory data science students in their first semester of R.

--> make sure this info is included somewhere:

From "Five Concrete Reasons Your Students Should Be Learning to Analyze Data in the Reproducible Paradigm" (@bray2014five)

* Helps with transparency: make clear data cleaning steps between raw data and final data
* Helps with collaboration: easier to share code when it takes very few steps to run on a different computer

The other big ideas on the importance/use of this package (not from sources):

* Make it easier for professors to grade students' code
* Should be usable by intro level data science students
* Possible use by reviewers of journal articles (and by those writing the articles)

# What Defines Reproducibility in Data Science?

# Creating Comprehensive Reproducibility Reports With "fertile"

Main points of different sources, as well as info about how they might be used for the paper.

Here are some sources we might use for motivation behind the project:

### The Reproducibility Crisis

Big idea: most scientific fields are facing a reproducibility crisis and poor statistical use is considered one of the important reasons behind this.

Not sure how useful this is due to the fact that it does not necessarily focus on the same kind of reproducibility we are looking at, which is code reproducibility rather than experimental reproducibility.

@baker20161

### Why is reproducibility important?

From Popper in `The Logic of Scientific Discovery`: "non-reproducible single occurrences are of no significance to science."

@popper2005logic

### What does reproducibility mean in data science?

"The ability of a researcher to duplicate the results of a prior study using the same materials as were used by the original investigator. That is, a second researcher might use the same raw data to build the same analysis files and implement the same statistical analysis in an attempt to yield the same results."

@goodman2016does

From "The Practice of Reproducibile Research": Can all of the figures/calculations related to the result be reproduced in a single button press or at least with a reasonably small effort?

@kitzes2017practice

### What makes a data science project reproducibile?

Another way to think about this is: what features does fertile have that match what different sources think needs to be checked to ensure reproducibility?

Some big ideas from the OpenSci discussion page:

* code should use tidy style
* project should have short vignette files -- most written code should be in an R/ directory
* code and data should be stored in separate folders
* a run environment and dependencies should be specified
* there should be a README file
* there should be a data dictionary
* one master script
* code should only use relative paths
* code should be well commented and all variables should be defined

Big ideas from "Packaging data analytical work reproducibly using R"

* research should be organized like an R package!
* clear separation of data, method, and output
* specify the computational environment that was used for the original analysis (typically in a plain text file)
* there should be a README describing the project and where to get started
* script files with reusable functions should go in an R/ directory
* raw data files should be kept in a data/ directory
* analysis scripts and report files should go in an analysis/ directory
* for simpe projects, scripts should be given ascending names. For more complicated ones, the use of some sort of makefile is recommended
* there should also be a DESCRIPTION file with information about the authors, project license, and software dependencies

@marwick2018packaging

From "The Practice of Reproducible Research"

* Are the data openly accessible? If hosted online, is the web address reliable long-term?
* Are they in a commonly used and well-documented file format? Avoid spreadsheets and instead use plain text data if possible!
* Is the raw data available? Is sufficient metadata provided?

* Are dependencies described properly?
* Is full history of source code available through a public version history

* Is there a README?
* Are functions documented?
* Is there narrative documentation explaining how the different pieces work together?
* Are there usage examples?

Folder setup might look like:

1. Raw Data
    + Data
    + README
2. Clean Data
    + Data
3. Results
    + Results file
4. Src
    + Analysis script
    + Script to clean data

@kitzes2017practice

From R OpenSci's Reproducibility Guide:

http://ropensci.github.io/reproducibility-guide/

* Is it clear where to begin?
* Can you determinine which files were used as input to create output files?
* Is there documentation about every result?
* Are exact versions of external applications noted?
* If using randomness, are seeds noted?
* Have you specified a license or noted licenses if you used other people's content?
* Are files easy to find?
* Is it clear what the most recent file is?
* Are there any folders that could be deleted?
* Is analysis output done hierarchically?
* Are there lots of manual data manipulation steps?

R OpenSci also has a discussion of code review practices where users commented. Not too much that was helpful from here (just a few odd comments):
https://discuss.ropensci.org/t/how-do-you-review-code-that-accompanies-a-research-project-or-paper-help-ropensci-plan-a-community-call/1321/43

From "A Guide to Reproducible Code in Ecology and Evolution" (these ideas are pretty universal, though):

A basic project structure:

* The data folder contains all input data (and metadata) used in the analysis.
* The doc folder contains the manuscript
* The figs directory contains figures generated by the analysis
* The output folder contains any type of intermediate or output files (e.g. simulation outputs, models, processed datasets, etc.). You might separate this and also have a cleaned-data folder.
* The R directory contains R scripts with function definitions.
* The reports folder contains RMarkdown files that document the analysis or report on results

* Consistent, ordered naming of scripts
* Use portable paths
* Write unit tests (only for advanced coding)
* Show the packages you used
* Record dependencies and versions of outside things you use

@cooper2017guide

# Karl Broman's Suggestions

* Encapsulate everything within one directory
* Separate raw data from derived data
* Separate data from code
* Use relative paths
* Choose filenames correctly
* Write README files

http://kbroman.org/steps2rr/pages/organize.html

# Why focus on R?

* R is the most popular language for statistical programming and is specifically designed for statistics
* R is great for reprodubility because the code is readable by users, and RMarkdown is a great way to show/explain processes
* R is easy to install and begin using

--> from https://openresearchsoftware.metajnl.com/articles/10.5334/jors.bu/print/

# Similar R Packages

### rrtools

https://github.com/benmarwick/rrtools

* Creates a basic R package named after your research topic
* Generates a license file
* Connects to GitHub and creates a repository
* Generates a README
* Generates a reproducible directory structure
* Creates a dockerfile
* Creates a minimal travis file
* Sets up testthat

### represtools

https://pirategrunt.com/represtools/

Has basic utility functions to support reproducibile research
• Gather
• Cook
• Analyze
• Present

### orderly

https://github.com/vimc/orderly

Focuses on automating a series of basic steps in the prodcess of writing analyses.
• Tracks inputs
• Stores multiple versions of an analysis (unclear whether this is different stages?)
• Track outputs
• Create analyses that depend on the outputs of previous analyses
• You have to declare what you're going to use and create BEFORE you start the analysis

### workflowr

https://f1000research.com/articles/8-1749

Aims to instill a particular workflow that helps make projects more reproducible.

• Version control (git)
• Literate programming (Rmd)
• Automatic checks & safeguards (seemingly related to whether different files in a project have been published on git, and not actually to do with code practices)
• Sharing code / results via browsable website

### checkers

https://github.com/ropenscilabs/checkers

• A framework for reviewing analysis projects
• Seems to be something run after-the-fact
• Currently a work in progress and not passing checks (as of Jan 2019)
• Can create custom checks
• Checks include:
 - Comments
 - Good coding style (lintr)
 - Use of version control
 - Whether preferred packages are used

# Msc sources to look at:

### Victoria Stodden's "Implementing Reproducible Research" book
https://books.google.com/books?hl=en&lr=&id=JcmSAwAAQBAJ&oi=fnd&pg=PP1&dq=Victoria+stodden+reproducibility&ots=ym_gvSxMNH&sig=B5N6FlkoocVjIprIZevtvrL33qQ#v=onepage&q=coombes&f=false

*Packaging data analytical work reproducibly using {R} (and friends) -- Marwick et al*

• The R package can be a suitable template for organizing files into a research compendium to enhance the reproducibility of research.

*Reproducible research and Biostatistics -- Peng*

• To meet the "reproducible" criterion in the biostatistics journal you must submit: a main script and any complementary side scripts, data and any other necessary files, a target file containing results.

*R OpenSci's suggestions:*

• To be reproducible, it should be easy to find files, be clear what each file is, and be easy to know what order to run files. There should be no extraneous files. Everything should be documented, dependencies noted, and seeds used for randomness. Automation should be used when possible.

*A guide to reproducible code in ecology and evolution -- Cooper et al*

• Reproducible work should include a clear and well-separated file system, well-named files, a list of package dependencies, good documentation. Preferably, RMarkdown and version control should be used.

*Karl Broman's suggestions:*

• Keep the project contained in one directory, separate the components (raw data, derived data, code), use relative paths, write a README, and name things clearly.

*What does research reproducibility mean? -- Goodman et al*

•The language and conceptual framework of research reproducibility vary across the sciences. There are no clear standards agreed upon by all scientific researchers.

*The practice of reproducible research: case studies and lessons from the data-intensive sciences -- Kitzes et al*

• Reproducibility examples from a variety of fields all vary significantly but they share key principles / practices in common: 1. Use clear separation, labeling, and documentation. 2. Automate documentation. 3. Design a workflow as a sequence of small steps glued together with intermediate outputs from one step feeding into the next.

>>>>

As argued by @Goodman341ps12, the language and conceptual framework of research reproducibility vary across the sciences. There are no clear standards agreed upon across fields.

The variety of recommendations and techniques for achieving reproducibility can be clearly seen in @kitzes2017practice, a collection of case studies on reproducibility practices from across the data-intensive sciences. Although the book idenfies no consensus on exact standards of reproducibility, several common trends and principles emerge from the case studies:

@@

@cooper2017guide, focused on data analysis in R, identifies similar important reproducibility components, reiterating the need to clearly labled, well-documented, and well-separated files. In addition, the paper recommends that a list of dependencies be published and version control be used.

@@

Much of the available literature focuses on file structure, organization, and naming, and **fertile`**s features are consistent with this. The *ideal* file structure is not agreed upon by academics, though there are some recommendations focused on R that are available. @marwick2018packaging provides the framework for file structure that **fertile** is based on: a structure similar to that of an R package, with an `R` folder, as well as `data`, `data-raw`, `inst` and `vignettes`.

@@

As we have seen, many existing R packages focusing on reproducibility are designed in a way such that they can only be used prior to the start of a project, often for the purposes of creating an organized file structure or ordered file hierarchy. 
**fertile**, on the other hand, can be used at any stage in the data analysis workflow. It is designed to be used on data analyses organized as R Projects (i.e. folders containing a `.Rproj` file). Once an R Project is created, **fertile** can provide benefits at any time, both while coding as well as after the fact.
**fertile** operates in two modes: proactively (to prevent reproducibility mistakes from happening in the first place), and retroactively (analyzing code that has already been written for potential problems). 
**fertile** is available for download at [https://github.com/baumer-lab/fertile](https://github.com/baumer-lab/fertile)

@@
